The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-09 rational.tex
_
nothing really
lots of stuff

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, ran
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a ``dual'' evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logical variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
{{<c,v,[]> |--> [c,v',[]]}} iff 
{{<c{dagger},v',[]> <--| [c{dagger},v,[]]}} 
\end{proposition}

@@

Consider the following two ways of purchasing an item that costs \$20.00:
\begin{enumerate}
\item You give the seller a \$20.00 bill.
\item You use a credit card to pay the seller.
\end{enumerate}
In both cases, the seller receives the money immediately but there is a
subtle difference. In the first transaction, the money received by the seller
corresponds to a value that has been produced earlier in the computation. In
the second transaction, the money may or may not exist yet: computationally,
a \emph{debt} is generated to compensate for the money received by the seller
and this debt travels ``backwards'' towards the bank where it is (hopefully)
reconciled.

The example suggests that the ability to consider values flowing backwards
enriches our computational model. This observation goes back to Filinski's
Masters thesis where continuations are identified with these negative values.
We discuss the connections to Filinski's work and others in more detail in
Sec.~\ref{sec:related}. For now, we note that in a conventional programming
language in which values can be copied and deleted at will, extreme care is
needed to keep track of negative values. Indeed, in the example above, if it
were possible to simply delete the variable corresponding to the debt, we
would have produced money out of nothing.

For this reason, our treatment of negative values in the context of
{{langRev}} is particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative values) will eventually have to match the negative
value with some corresponding value, effectively using the isomorphism in the
right-to-left direction. In more detail, we can model the credit card
transaction above using the following program (shown diagrammatically):

\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

\roshan{I chose to draw the figure with b instead of a, since b is what
  we use for the types. Should this be a? Also I chose to make the
  swap and 0 wires, explicit.Is this what you intended? }

CLEAN UP the following based on the figure. In contrast, in our setting, one
can start from the empty type $0$, introduce a positive value and its
negative counterpart, and let each of these flow in arbitrary ways. The
entire framework guarantees that neither the value nor its negative
counterpart will be deleted or duplicated and hence that, in any closed
program, the ``debt'' corresponding to the negative value is paid off exactly
once. Computationally we model the first transaction using essentially the
identity function which receives a \$20.00 bill as input from the buyer and
propagates it on its output to the seller. The second transaction is more
complicated. We model it as shown in the circuit below: There are two ways to
understand this circuit that are both quite instructive. Let's first examine
the type structure of each of the combinators that comprise the circuit. The
first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
produced from nothing so to speak by generating an equivalent debt that
travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
the execution. It goes forwards in time so to speak, comes back, and then goes
forwards again.

It is critical that the framework in which the negative types are introduced
is a framework in which all information is preserved, with no duplication or
erasure. This guarantees that the generated debts are paid once and exactly
once.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
defined only if {{b1}} and {{b2}} are entirely positive types. This is
similar to the constraint that Zeilberger imposes to explain intuitionistic
polarity and delimited control~\cite{10.1109/LICS.2010.23}.

@@

Additive duality is added to {{langRev}} by addition of negative types
denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
extend the language as follows:

\noindent
with the types judgments

%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b

We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
dually on the right is {{eps+}} showing the map from {{-b+b}} to
0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can
implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

The usual interpretation of the type {{b1+b2}} is that we either have
an appropriately tagged value of type {{b1}} or an appropriately
tagged value of type {{b2}}. This interpretation is maintained in the
presence of {{eta+}} and {{eps+}} in the following sense: a value of
type {{right v : -b + b}} flowing into an {{eps+}} on the lower wire
switches into a value {{left (-v): -b + b}} that flows backwards on the
upper wire. The inversion of direction is captured by the negative
sign on the value.

@@

The type {{a/b}} is prominent in the Lambek-Grishin calculus which is
extensively used in computational linguistics. In the common interpretation,
a value of type {{a/b}} is a value of some type $c$ such that when put in a
context of type $b$, the result is a value of type $a$. For example, assuming
contexts are represented as functions, a value of type $\texttt{bool} /
(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
type $\texttt{bool}$.

In our case, the situation is simpler. If the goal is to produce a value of
type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
have type {{(a * b) / b}}. In most settings, this type would be equivalent to
{{a}}. However in our setting, the constraint that this value must eventually
fit in a larger value that supplies the {{b}} is explicitly recorded and must
be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
which when used in the left-to-right direction allows the creation of a value
and a corresponding constraint on the context.  Both the value and the
constraint can flow in arbitrary ways during the computation but eventually
in a closed program with no ``dangling'' constraints, the isomorphism should
be used in the right-to-left direction to resolve the constraints.  To
understand the computational interpretation, consider the following example.

The goal is to produce a value of type $(\texttt{bool} \times
\texttt{int})$. One part of the program, $e_1$, knows how to produce the
value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
to produce the value of type $\texttt{bool}$ (say true). Computationally, we
model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
the two values meet, we can group the components as follows:
\[
(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
\]
Using the isomorphism in the right-to-left direction on the last two pairs,
forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
with the pair $(true,3)$ as desired.

To summarize, one can think of a value of type $a/b$ as a value that imposes
a constraint on its context of use: it is a value that requires its context
to be of type $b$ and only if that condition is satisfied, can the value be
considered as a value of type $a$. In the simplest case, the type $1/b$ just
constraints the context to be of type $b$. By allowing the fractional types
to be first-class, we can separate the generation of constraints from their
use. Both the value and the constraint can flow in arbitrary ways during the
computation but eventually in a closed program with no ``dangling''
constraints, the isomorphism should be used in the right-to-left direction to
resolve the constraints.

NEED to work out the example in detail (or perhaps another example that's
better.

Again it is critical that the framework in which the fractional types are
introduced is a framework in which all information is preserved, with no
duplication or erasure. This guarantees that the generated constraints are
satisfied once and exactly once.

\begin{center}
  \includegraphics{diagrams/algebra2.pdf}
\end{center}

@@

Multiplicative duality is added to {{langRev}} by adding types
{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
{{a/b}}.

%subcode{bnf} include main
% Value Types, b = ... | inv(b)
% Values, v = ... | inv(v)
%
% Isomorphisms, iso &=& ... | eta* | eps*

@@

%subcode{opsem} include main
% eta* &: 1 <-> inv(b) * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- inv(v) : inv(b)

As with additive duals, we represent {{eta*}} and {{eps*}} using
`U'-shaped connectors. In the general case the wires for {{1}} will be
dropped as it can freely introduced or eliminated using {{uniti}} and
{{unite}}.

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{inv(b)}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{inv(b)}} is
always the dual of the chosen one.

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{inv(b)}}. These two values are free to move independently
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

@@

\begin{enumerate}
\item {{eta* s () |--> ((inv(v), v), s)}}
  where {{v}} is a fresh logic variable.

@@

MUST TALK ABOUT AND CITE ABRAMSKY AND COECKE'S TR ABOUT QUANTUM
COMPUTING.

These constructions are applicable to the additive and the
multiplicative monoid.

\paragraph*{Functions}

\paragraph*{Trace }

\paragraph*{Principium Contradictiones}

@@

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logical variables:
they can be created and instantiated later once their context of use is
determined. Although a formal connection is lacking, it is clear that, at an
intuitive level, acceptors are entities that combine elements of our negative
and fractional types.

@@

%% Jacques

>>>>

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, ran
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a ``dual'' evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
{{<c,v,[]> |--> [c,v',[]]}} iff 
{{<c{dagger},v',[]> |--> [c{dagger},v,[]]}} 
\end{proposition}

@@

The structure of this section and the next are the same. The general outline
is as follows:\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}. As will be detailed in this
section and the next, the compact closed structure provides several
properties of interest: (i) morphisms or wires are allowed to run from right
to left, (ii) the structure admits a trace that, depending on the situation,
can be used to implement various notions of loops and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual operator is
an involution, and (iv) the structure is equipped with objects representing
higher-order functions and a morphism \textit{eval} that applies these
functional objects. Interestingly each monoidal structure provides the same
ingredients, resulting in two dualities, two traces, two involutions, and two
notions of higher-order functions.

Describing the syntax and types of an extension to {{langRev}} with additive
duality is fairly straightforward. Basically, we extend the language with
negative types denoted {{-b}}, negative values denoted {{-v}}, and two
isomorphisms {{eta+}} and {{eps+}}:

The isomorphisms {{eta+}} and {{eps+}} witness the following
isomorphism:
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+
And as expected we have the following typing rule for negative value:
%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b

For the graphical language, we visually present {{eta+}} and {{eps+}} as
U-shaped connectors. On the left below is {{eta+}} showing the map from {{0}}
to {{-b+b}} and dually on the right is {{eps+}} showing the map from {{-b+b}}
to 0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can implicitly
introduce/eliminate then using {{zeroi}}/{{zeroe}}:
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

The usual interpretation of the type {{b1+b2}} is that we either have an
appropriately tagged value of type {{b1}} or an appropriately tagged value of
type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
flowing into an {{eps+}} on the lower wire switches to a value 
{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
direction is captured by the negative sign on the value. 
As an example, the diagram representing the isomorphism from the introduction
is:
\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compact Closure with Additive Inverses}

\paragraph*{Trace}

\paragraph*{Involution (Principium Contradictiones)}

\paragraph*{Higher-Order Functions}

%%%%%%%%%%%%%

For this reason, our treatment of negative values in the context of
{{langRev}} is particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative values) will eventually have to match the negative
value with some corresponding value, effectively using the isomorphism in the
right-to-left direction.

\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
defined only if {{b1}} and {{b2}} are entirely positive types. This is
similar to the constraint that Zeilberger imposes to explain intuitionistic
polarity and delimited control~\cite{10.1109/LICS.2010.23}.

@@

@@

The section follows the outline of the previous one, first presenting the new
syntax, and discussing the graphical language and the properites implied by
the underlying categorical structure.

@@

Multiplicative duality is added to {{langRev}} by adding types
{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.

%subcode{bnf} include main
% Value Types, b = ... | 1/b
% Values, v = ... | 1/v
%
% Isomorphisms, iso &=& ... | eta* | eps*

@@

%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- 1/v : 1/b

As with additive duals, we represent {{eta*}} and {{eps*}} using
U-shaped connectors. In the general case the wires for {{1}} will be
dropped as it can freely introduced or eliminated using {{uniti}} and
{{unite}}.

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{1/b}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{1/b}} is
always the dual of the chosen one.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compact Closure with Multiplicative Inverses}

The constructions for functions, involution, and trace are identical,
replacing {{+}} by {{*}} mutatis mutandis.

%%%%%

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{1/b}}. These two values are free to move independently
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

once we combine the two structures, we seem to retain the monoidal
structure. How!?

@@

\begin{enumerate}
\item {{eta* s () |--> ((1/v, v), s)}}
  where {{v}} is a fresh logic variable.

@@

@@

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logic variables:
they can be created and instantiated later once their context of use is
determined. Although a formal connection is lacking, it is clear that, at an
intuitive level, acceptors are entities that combine elements of our negative
and fractional types.

@@

%% Jacques Carette, Michael Adams
